---
title: "Building Agents"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building Agents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

In **aisdk**, an **Agent** is a specialized AI worker. While a `ChatSession` holds conversation state, an `Agent` holds **capability**: a persona (system prompt), a set of tools, and optionally, a specialized library of skills.

## The Agent Class

Agents are implemented as R6 classes. You can create one using the `Agent$new()` constructor or the `create_agent()` factory function.

```{r}
library(aisdk)

# Create a specialized Research Agent
researcher <- create_agent(
  name = "Researcher",
  description = "Finds and summarizes information on specific topics",
  system_prompt = "You are a professional researcher. Provide citations for your findings."
)

print(researcher)
```

## Running an Agent

Use the `$run()` method to give the agent a task.

```{r}
model <- create_openai()$language_model("gpt-4o")

result <- researcher$run(
  task = "Summarize the latest trends in R package development for 2025.",
  model = model
)

cat(result$text)
```

## Agent Library

The `aisdk` package includes a library of pre-configured specialists. These are optimized for common data science tasks.

```{r}
# A Coder agent with a built-in R execution sandbox
coder <- create_coder_agent()

# A Visualizer agent optimized for ggplot2
visualizer <- create_visualizer_agent()

# A Planner agent for breaking down complex objectives
planner <- create_planner_agent()
```

## Agent Autonomy and Safety

When agents execute code dynamically (e.g., using the Coder agent's sandbox), **aisdk** provides a sophisticated security model to keep your environment safe:

1. **AST Static Analysis**: Before any LLM-generated R code is executed, `check_ast_safety()` analyzes the Abstract Syntax Tree. It silently blocks forbidden functions (like `system()`) and flags potentially risky ones (like `unlink()`).
2. **Variable Registry**: Protect critical session variables from being shadowed, modified, or heavily copied by the agent. Use `sdk_protect_var("my_important_df", locked = TRUE, cost = "High")`.
3. **Human-in-the-Loop (HITL)**: For "YELLOW" risk operations (like writing files or modifying the global environment), `request_authorization()` automatically pauses execution and prompts the user for explicit permission via the console before proceeding.

## Multi-Agent Orchestration

Agents can work together using an `AgentRegistry` and a `Flow`. A "Manager" agent can delegate tasks to "Worker" agents based on their descriptions.

```{r}
# 1. Register specialists
registry <- create_agent_registry(list(coder, visualizer))

# 2. Create a Flow with a shared session
session <- ChatSession$new(model = model)
flow <- Flow$new(session = session, registry = registry)

# 3. Create a Manager agent
manager <- create_agent(
  name = "Manager",
  description = "Orchestrates data analysis tasks",
  system_prompt = "You are a project manager. Delegate tasks to Coder or Visualizer as needed."
)

# 4. Run the coordinated flow
# The Manager will automatically get tools like 'delegate_to_CoderAgent'
result <- flow$run(
  agent = manager,
  task = "Load the 'iris' dataset, calculate means by species, and plot the result."
)
```

## Low-Level Control

You can manually convert an agent into a tool that another agent (or `generate_text`) can call:

```{r}
coder_tool <- coder$as_tool()

response <- generate_text(
  model = model,
  prompt = "Write a function to calculate Fibonacci numbers.",
  tools = list(coder_tool)
)
```

For more advanced agent behaviors, see the [Skills](skills.html) vignette.
