---
title: "Console Chat"
format: html
vignette: >
  %\VignetteIndexEntry{Console Chat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    eval: false
---

`console_chat()` is the terminal-first interactive interface in **aisdk**. It
wraps a `ChatSession` in a REPL, adds an agent-oriented command surface, and
provides a structured console UI with status, inspection, and debugging
layers.

This article focuses on how to use the console effectively today: how to start
it, how the three view modes differ, how the inspector overlay works, and which
slash commands matter in day-to-day work.

## Why Use `console_chat()`

`console_chat()` is useful when you want:

- a fast terminal workflow instead of a Shiny or document-based interface
- a persistent chat session with model switching and save/load support
- an agent that can execute R code, shell commands, and file operations
- compact default output that still lets you inspect tool-heavy turns
- a troubleshooting path that stays inside the console

It is especially effective for iterative analysis, quick automation, data
inspection, and prompt-tool experimentation where leaving the terminal would
slow you down.

## Starting the Console

### Default agent mode

The default startup path enables the built-in terminal agent:

```{r}
# Start the default console agent
# console_chat("openai:gpt-4o")
```

In this mode, the console can call tools for:

- shell execution
- reading and writing files
- sandboxed R code execution
- local R execution when explicitly enabled

### Simple chat mode

If you want plain chat without agent tools:

```{r}
# Disable agent tools
# console_chat("openai:gpt-4o", agent = NULL)
```

### Debug startup

If you are developing tools, providers, or console behavior itself:

```{r}
# Start directly in debug mode
# console_chat("openai:gpt-4o", verbose = TRUE)
```

### Session and model inputs

`console_chat()` accepts:

- a model string, such as `"openai:gpt-4o"`
- a `LanguageModelV1` object
- an existing `ChatSession`

That means you can either launch from scratch or drop into an existing session
that already contains history, memory, or a custom agent.

## UI Model

The console UI is no longer just a print stream. It currently uses a shared
frame builder that assembles several surfaces before rendering them through a
conservative append-only terminal path.

### Status bar

The status bar stays visible as the top summary surface and shows:

- current model
- sandbox mode
- current view mode
- streaming on/off
- local execution on/off
- current tool state

This gives you quick context without inspecting the transcript.

### Tool timeline

In inspect mode, tool-heavy turns emit a compact timeline summary rather than
raw payload dumps. Each line records:

- tool index
- tool name
- tool status
- argument summary
- result summary
- message/warning counts when present
- elapsed time when available

### Inspector overlay

The inspector overlay is a separate surface backed by real overlay state. It
can open for:

- the latest turn
- a specific tool inside the latest turn

It currently renders as a boxed overlay section below the status bar, but it is
already modeled as a first-class overlay rather than an ad hoc print block.

## View Modes

`console_chat()` currently supports three presentation modes.

### `clean`

This is the default mode.

Use it when you want:

- calm transcript output
- minimal tool noise
- no raw tool payloads
- no thinking output

This is the right mode for most day-to-day usage.

### `inspect`

This is the middle layer between normal use and full debugging.

Use it when you want:

- compact transcript output
- per-turn tool timeline summaries
- access to the inspector overlay
- warnings/messages preserved without raw log spam

This is the recommended mode whenever a turn used tools and you need to answer
"what happened?"

### `debug`

This is the troubleshooting mode.

Use it when you want:

- detailed tool logging
- thinking visibility
- lower-level execution visibility
- development-oriented introspection

This mode is intended for package development and difficult failures, not as
the normal default.

## Slash Commands

### Session control

- `/help`: show the command list
- `/quit`, `/exit`: end the console session
- `/save [path]`: save the current session
- `/load <path>`: restore a saved session
- `/clear`: clear conversation history

### Model and runtime control

- `/model <id>`: switch the model without losing the session
- `/stream [on|off]`: toggle token streaming
- `/local [on|off]`: enable or disable local execution mode
- `/debug [on|off]`: switch between clean and debug behavior
- `/inspect [on|off]`: switch into or out of inspect mode

### Visibility and inspection

- `/history`: show session history
- `/stats`: show token and tool usage statistics
- `/inspect turn`: open the inspector overlay for the latest turn
- `/inspect tool <index>`: open the overlay for a specific tool
- `/inspect next`: move to the next tool in the active inspector context
- `/inspect prev`: move to the previous tool in the active inspector context
- `/inspect close`: close the active inspector overlay

## Inspector Workflow

The most useful console workflow today is:

1. Stay in `clean` mode while chatting normally.
2. Switch to `inspect` after a tool-heavy turn.
3. Review the timeline summary.
4. Open the latest turn in the inspector overlay.
5. Jump into a specific tool if needed.
6. Navigate between tools with `/inspect next` and `/inspect prev`.
7. Close the overlay when done.

### What the inspector shows

At turn level, the overlay shows:

- turn id
- current phase
- elapsed time
- compact previews of user and assistant content
- collected messages and warnings
- timeline entries for all tools in the turn

At tool level, the overlay shows:

- tool name and status
- elapsed time
- argument summary
- result summary
- messages and warnings
- compact raw previews of arguments and results

This preserves a useful hierarchy:

- `clean` keeps the transcript calm
- `inspect` shows what executed
- the overlay shows enough detail to diagnose a turn
- `debug` remains available when you need the full machinery

## Agent Capabilities

In default agent mode, `console_chat()` is not just a chat box. It is a
terminal agent surface.

Typical tasks include:

- asking the model to inspect files in a project
- running small R analyses
- checking package versions or environment details
- reading data files
- writing small helper files
- running shell commands inside the configured working directory
- analyzing screenshots, charts, or figures when the user asks visual questions
- generating image assets when the user asks for posters, hero images, or illustrations
- editing the most recently generated image when the user asks for follow-up changes

The console keeps these actions compact by summarizing tool progress inline and
moving deeper detail into the inspect layer.

## Native Image Workflows

Image and vision support in `console_chat()` is intended to behave like a native
agent capability rather than a separate command language.

That means the recommended user experience is:

- the user asks naturally for screenshot analysis, OCR, chart reading, or UI critique
- the agent decides to use image-analysis tools on its own
- the user asks for a poster, thumbnail, product shot, or illustration
- the agent decides to call image-generation tools on its own
- the user says \"make the last one blue\" or \"edit the previous render\"
- the agent reuses the most recent remembered image artifact when possible

The current console agent has dedicated image tools behind the scenes for:

- local image discovery
- image analysis
- structured extraction from images
- image generation
- image editing
- recalling recent generated or edited image artifacts

This keeps image work aligned with the rest of the agent workflow: the user asks
for an outcome, and the agent chooses the right tool path.

In the current design, the agent should usually try this order on its own:

1. reuse the most recent relevant image artifact if one exists
2. search the working directory for likely image candidates
3. ask the user for a path only if neither step produced a clear candidate

The console agent can also work across multiple images when that is the most
natural interpretation of the request, for example:

- comparing several screenshots
- extracting repeated fields from multiple forms
- reviewing a small batch of generated assets

This is intentionally designed as an agent behavior rather than a separate
special command language.

## Persistence and Session Continuity

Because `console_chat()` sits on top of `ChatSession`, it inherits the session
model rather than inventing a separate console-only state system.

That means you can:

- switch models mid-session
- save and reload sessions
- preserve history across runs
- retain access to shared session memory and environment-backed state

Typical pattern:

```{r}
# Save before stopping work
# /save analysis_session.rds

# Restore later
# /load analysis_session.rds
```

## Local Execution and Sandbox Behavior

The status bar makes execution context visible, which matters when an agent can
touch the local environment.

Key points:

- the console starts with the configured sandbox mode
- local execution is separately visible and separately toggleable
- you can inspect the current runtime state from the status bar instead of
  guessing

When you enable local execution, the agent can work in the shared session
environment more directly, so this should be treated as a deliberate workflow
choice rather than a background detail.

## Recommended Usage Pattern

For most users, the practical workflow is:

1. Start in default agent mode.
2. Work in `clean` mode for normal conversation.
3. Switch to `inspect` when a turn executed tools.
4. Open the inspector overlay only when you need detail.
5. Use `debug` when troubleshooting or developing console behavior.
6. Save the session when you want to continue later.

This keeps the console productive in normal use without giving up the ability
to inspect execution deeply.

## Current Limitations

The current console is substantially more structured than the original REPL,
but it is still mid-transition toward a fuller TUI architecture.

Current limitations to be aware of:

- overlay rendering still uses an append-only terminal path
- frame building is in place, but region ownership is still evolving
- keyboard-driven focus navigation is not yet a complete alternative to slash
  commands
- the composer is still based on the existing input path rather than a fully
  custom multiline editor

These constraints do not block normal use, but they explain why the inspector
is command-driven today and why overlay behavior is intentionally conservative.

## Related Articles

- [Get Started](get_started.html): introductory package overview
- [Chat Sessions](sessions.html): state, persistence, and shared environment
- [Building Agents](agents.html): how agent tools and sessions fit together
- [Computer Abstraction Layer](computer_abstraction.html): deeper background on
  the terminal-oriented tool surface
