---
title: "Get Started with aisdk"
format: html
vignette: >
  %\VignetteIndexEntry{Get Started with aisdk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    eval: false
---

**aisdk** is a production-grade AI SDK for R, providing a unified interface for multiple AI model providers (OpenAI, Anthropic, Gemini, DeepSeek, etc.). It features a layered architecture, robust error handling, and advanced capabilities like multi-agent orchestration and a distributed MCP (Model Context Protocol) ecosystem.

## Installation

You can install the development version from GitHub:

```{r}
# install.packages("devtools")
devtools::install_github("YuLab-SMU/aisdk")
```

## Basic Usage

### 1. Configuration

Load the library and set your API keys as environment variables. We recommend using an `.Renviron` file for persistence.

```{r}
library(aisdk)

# Set API keys (Example)
# Sys.setenv(OPENAI_API_KEY = "sk-...")
# Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
```

### 2. Creating a Model

Create a model instance using provider factory functions.

```{r}
# Create an OpenAI model
model <- create_openai()$language_model("gpt-4o")

# Create an Anthropic / Gemini model
# model <- create_anthropic()$language_model("claude-3-5-sonnet-20241022")
# model <- create_gemini()$language_model("gemini-1.5-pro")

# Create models from other supported providers (DeepSeek, Volcengine, Stepfun, etc.)
# model <- create_deepseek()$language_model("deepseek-chat")
# model <- create_volcengine()$language_model("your-endpoint-id")
# model <- create_stepfun()$language_model("step-1-8k")
```

If you want a package-wide default, set it once and omit `model` from
high-level helpers:

```{r}
set_model("openai:gpt-4o")

# Equivalent shortcut
model("openai:gpt-4o")

# Inspect the current default
model()
```

### 3. Generating Text

Use `generate_text()` for simple request-response interactions.

```{r}
# Simple prompt
response <- generate_text(
  prompt = "What are the advantages of using R6 classes in R?"
)

cat(response$text)

# Using a system prompt for behavior control
response <- generate_text(
  model = model,
  system = "You are a helpful R programming expert.",
  prompt = "Explain the difference between `lapply` and `sapply`."
)
```

### 4. Streaming Responses

For real-time output, use `stream_text()` with a callback.

```{r}
stream_text(
  prompt = "Write a short poem about data science.",
  callback = function(chunk, done) {
    if (!done) cat(chunk)
  }
)
```

### 5. Managing State with ChatSession

For stateful conversations, use `ChatSession`. It automatically manages history and provides a higher-level API.

```{r}
# Initialize a session
session <- ChatSession$new()

# First message
res1 <- session$send("Hi, I'm analyzing gene expression data.")
cat(res1$text)

# Second message (context is preserved)
res2 <- session$send("What R packages would you recommend for normalization?")
cat(res2$text)

# Check history
# session$history()
```

### 6. Tools and Function Calling

**aisdk** makes it easy to give LLMs access to R functions.

```{r}
# Define a tool using the schema DSL
weather_tool <- tool(
  name = "get_weather",
  description = "Get current weather for a city",
  parameters = z_object(
    city = z_string("The city name, e.g., Beijing")
  ),
  execute = function(args) {
    paste("The weather in", args$city, "is sunny, 25 degrees C.")
  }
)

# Use the tool
response <- generate_text(
  model = model,
  prompt = "What's the weather like in Shanghai?",
  tools = list(weather_tool),
  max_steps = 3
)

cat(response$text)
```

### 7. Interactive Console Chat

For a more interactive experience, **aisdk** provides a built-in REPL (Read-Eval-Print Loop) called `console_chat()`. This allows you to chat with the model directly in your R terminal, with support for streaming, slash commands, and an intelligent agent that can execute R code or bash commands.

```{r}
# Optional: set a package-wide default first
# set_model("openai:gpt-4o")

# Start an interactive chat with an intelligent agent (default)
# console_chat("openai:gpt-4o")

# Start a simple chat without agent tools
# console_chat("openai:gpt-4o", agent = NULL)
```

`console_chat()` now exposes a structured terminal workflow rather than a plain print stream:

- A persistent status bar shows the current model, sandbox mode, view mode, streaming state, local execution state, and tool activity.
- Agent mode can run shell commands, read and write files, and execute R code through natural language.
- Tool-heavy turns stay compact in normal use, but remain inspectable when you need to understand what happened.
- The console keeps a shared frame structure for status, timeline, and overlay surfaces, while still degrading safely to an append-only terminal path.

#### View modes

- `clean`: the default mode. Keeps the transcript calm and hides raw tool payloads.
- `inspect`: adds per-turn tool timeline summaries and an inspector overlay for drilling into the latest turn or an individual tool.
- `debug`: enables detailed tool logging and thinking output for development and troubleshooting.

You can switch modes live:

```{r}
# Enter inspect mode
# /inspect on

# Enter debug mode
# /debug on

# Return to compact clean mode
# /debug off
```

#### Inspector workflow

The inspector is useful when a turn used tools and you want to understand the execution path without switching the entire console into debug mode.

- `/inspect turn`: opens the inspector overlay for the latest turn.
- `/inspect tool <index>`: opens the inspector overlay for a specific tool in the latest turn.
- `/inspect next` / `/inspect prev`: moves between tools inside the current inspector context.
- `/inspect close`: closes the active inspector overlay.

The inspector overlay shows:

- turn or tool identity
- elapsed time
- summary of arguments and results
- captured messages and warnings
- compact previews of raw arguments and raw results

#### Common slash commands

- `/help`: show all available commands.
- `/model <id>`: switch the model on the fly.
- `/save [path]` / `/load <path>`: persist and restore sessions.
- `/history`: inspect conversation history.
- `/stats`: inspect session usage statistics.
- `/stream [on|off]`: toggle token streaming.
- `/local [on|off]`: toggle local execution mode in the shared session environment.
- `/clear`: reset the conversation history.
- `/quit`: exit the interactive session.

#### Practical usage pattern

In day-to-day work, the recommended workflow is:

1. Start in `clean` mode for normal conversation.
2. Switch to `inspect` when a turn uses tools and you want a concise explanation of what ran.
3. Open the inspector overlay for the latest turn or a specific tool.
4. Use `debug` only when developing providers, tools, or console behavior itself.

## Next Steps

Explore the following vignettes for more advanced features:

- [Providers](providers.html): Configure models, connect via Custom Providers, and use AIHubMix advanced compatibility wrappers.
- [Agents](agents.html): Build specialized AI workers and multi-agent systems.
- [Tools](tools.html): Deep dive into the Tool system and Schema DSL.
- [Sessions](sessions.html): Advanced session management and persistence.
- [Console Chat](console_chat.html): Full guide to the terminal REPL, inspect/debug modes, and inspector overlay workflow.
- [MCP](mcp.html): Connect to external tools via the Model Context Protocol.
- [Skills](skills.html): Progressive knowledge loading for agents.
