---
title: "Get Started with aisdk"
format: html
vignette: >
  %\VignetteIndexEntry{Get Started with aisdk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    eval: false
---

**aisdk** is a production-grade AI SDK for R, providing a unified interface for multiple AI model providers (OpenAI, Anthropic, Gemini, DeepSeek, etc.). It features a layered architecture, robust error handling, and advanced capabilities like multi-agent orchestration and a distributed MCP (Model Context Protocol) ecosystem.

## Installation

You can install the development version from GitHub:

```{r}
# install.packages("devtools")
devtools::install_github("YuLab-SMU/aisdk")
```

## Basic Usage

### 1. Configuration

Load the library and set your API keys as environment variables. We recommend using an `.Renviron` file for persistence.

```{r}
library(aisdk)

# Set API keys (Example)
# Sys.setenv(OPENAI_API_KEY = "sk-...")
# Sys.setenv(ANTHROPIC_API_KEY = "sk-ant-...")
```

### 2. Creating a Model

Create a model instance using provider factory functions.

```{r}
# Create an OpenAI model
model <- create_openai()$language_model("gpt-4o")

# Create an Anthropic / Gemini model
# model <- create_anthropic()$language_model("claude-3-5-sonnet-20241022")
# model <- create_gemini()$language_model("gemini-1.5-pro")

# Create models from other supported providers (DeepSeek, Volcengine, Stepfun, etc.)
# model <- create_deepseek()$language_model("deepseek-chat")
# model <- create_volcengine()$language_model("your-endpoint-id")
# model <- create_stepfun()$language_model("step-1-8k")
```

### 3. Generating Text

Use `generate_text()` for simple request-response interactions.

```{r}
# Simple prompt
response <- generate_text(
  model = model,
  prompt = "What are the advantages of using R6 classes in R?"
)

cat(response$text)

# Using a system prompt for behavior control
response <- generate_text(
  model = model,
  system = "You are a helpful R programming expert.",
  prompt = "Explain the difference between `lapply` and `sapply`."
)
```

### 4. Streaming Responses

For real-time output, use `stream_text()` with a callback.

```{r}
stream_text(
  model = model,
  prompt = "Write a short poem about data science.",
  callback = function(chunk, done) {
    if (!done) cat(chunk)
  }
)
```

### 5. Managing State with ChatSession

For stateful conversations, use `ChatSession`. It automatically manages history and provides a higher-level API.

```{r}
# Initialize a session
session <- ChatSession$new(model = model)

# First message
res1 <- session$send("Hi, I'm analyzing gene expression data.")
cat(res1$text)

# Second message (context is preserved)
res2 <- session$send("What R packages would you recommend for normalization?")
cat(res2$text)

# Check history
# session$history()
```

### 6. Tools and Function Calling

**aisdk** makes it easy to give LLMs access to R functions.

```{r}
# Define a tool using the schema DSL
weather_tool <- tool(
  name = "get_weather",
  description = "Get current weather for a city",
  parameters = z_object(
    city = z_string("The city name, e.g., Beijing")
  ),
  execute = function(args) {
    paste("The weather in", args$city, "is sunny, 25°C.")
  }
)

# Use the tool
response <- generate_text(
  model = model,
  prompt = "What's the weather like in Shanghai?",
  tools = list(weather_tool),
  max_steps = 3
)

cat(response$text)
```

## Next Steps

Explore the following vignettes for more advanced features:

- [Agents](agents.html): Build specialized AI workers and multi-agent systems.
- [Tools](tools.html): Deep dive into the Tool system and Schema DSL.
- [Sessions](sessions.html): Advanced session management and persistence.
- [MCP](mcp.html): Connect to external tools via the Model Context Protocol.
- [Skills](skills.html): Progressive knowledge loading for agents.
