---
title: "Shiny Integration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

**aisdk** provides seamless integration with Shiny, allowing you to build AI-powered applications where the AI can control the UI and interact with reactive input/output.

## The AI Chat UI

Use `aiChatUI()` and `aiChatServer()` to add a chat interface.

```r
library(shiny)
library(aisdk)

ui <- fluidPage(
  aiChatUI("chat")
)

server <- function(input, output, session) {
  aiChatServer("chat", model = "openai:gpt-4o")
}

shinyApp(ui, server)
```

## AI-Controlled Apps (The Copilot Pattern)

You can give the AI tools to modify `reactiveValues`, effectively letting it control your app.

### 1. Define Reactive Values
```r
rv <- reactiveValues(
  plot_color = "blue",
  title = "My Plot"
)
```

### 2. Create Reactive Tools
Use `reactive_tool()` to define tools that can read/write these values.

```r
change_color <- reactive_tool(
  name = "change_color",
  description = "Change the plot color",
  parameters = z_object(color = z_string()),
  execute = function(rv, session, color) {
    rv$plot_color <- color
    paste("Changed color to", color)
  }
)
```

### 3. Initialize Server with Tools
```r
server <- function(input, output, session) {
  # ... rv definition ...
  
  # Wrap tools with current session context
  tools <- wrap_reactive_tools(list(change_color), rv, session)
  
  # Provide context so AI "sees" the current state
  get_context <- reactive({
    list(current_color = rv$plot_color)
  })

  aiChatServer(
    "chat",
    model = "openai:gpt-4o",
    tools = tools,
    context = get_context
  )
}
```

Now the user can say "Change the plot to red", and the AI will execute the tool, updating the Shiny app instantly.

## RMarkdown and Quarto Integration

**aisdk** also includes a custom knitr engine `{ai}` that allows you to use LLMs to generate and execute R code directly within RMarkdown or Quarto documents. 

### 1. Register the AI Engine
Before using `{ai}` chunks in your document, you need to register the engine. It's recommended to do this in an initial setup chunk.

```r
library(aisdk)
register_ai_engine()
```

### 2. Basic Usage
Once registered, you can use `{ai}` as the chunk language. You simply provide instructions or prompts within the chunk.

````markdown
```{ai}
Create a scatter plot of mpg vs wt from the mtcars dataset, 
colored by the number of cylinders (cyl). Make it look nice using ggplot2.
```
````

Behind the scenes, the engine will:
1. Send your prompt to the LLM.
2. Extract the R code the LLM generates.
3. Automatically execute the code (if `eval = TRUE`) in the document's environment.
4. Output the LLM's explanation, the generated code (if `echo = TRUE`), and the code's output (plots, tables, console text).

### 3. Chunk Options

The `{ai}` engine supports several custom chunk options:

- **`model`**: Specifies the LLM to use (e.g., `model = "anthropic:claude-3-5-sonnet-20241022"`). Defaults to the value of `getOption("aisdk.default_model")`.
- **`session`**: A character string naming the chat session. The engine maintains session history across multiple `{ai}` chunks within the same document, allowing you to ask follow-up questions. Defaults to `"default"`.
- **`new_session`**: If `TRUE`, clears the previous history for the specified session and starts fresh.
- **`system`**: Custom system prompt instructions for the LLM.
- **`context`**: Controls how data is passed from your R environment to the LLM.

#### Context Auto-Detection
The most powerful feature of the `{ai}` engine is **context**. If your prompt mentions variables that exist in the knitr environment, the engine will automatically detect them and send summaries of their structure/data to the LLM. 

For example, if you load a dataset in a previous `{r}` chunk:

````markdown
```{r}
my_data <- data.frame(category = c("A", "A", "B", "C"), value = rnorm(4))
```

```{ai}
Calculate the mean of 'value' grouped by 'category' in my_data using dplyr.
```
````
The engine will see `my_data` in your prompt, find it in the environment, and send its schema to the LLM so it writes accurate code.

You can manually control this with the `context` chunk option:
- `context = "my_data"`: Explicitly include only `my_data`.
- `context = c("data1", "data2")`: Include multiple specific objects.
- `context = FALSE`: Disable context passing entirely.
