---
title: "Connectivity (MCP)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Connectivity (MCP)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The **Model Context Protocol (MCP)** is an open standard that enables AI models to connect to external data sources and tools. `aisdk` provides a full-featured MCP client that can bridge any MCP server into your R-based agents.

## Why MCP?

Instead of writing custom R wrappers for every API, you can use existing MCP servers (available in Node.js, Python, and Go) to give your agents access to:
- **Filesystems**: Local or cloud storage.
- **APIs**: GitHub, Slack, Google Drive, etc.
- **Databases**: Postgres, SQLite, MongoDB.
- **Search**: Brave Search, Google Search.

## Using the MCP Client

The `McpClient` class manages the connection to an external MCP server process via `stdio`.

### 1. Connecting to a Server

You can start a server process using `create_mcp_client()`.

```{r}
library(aisdk)

# Example: Connect to the SQLite MCP server
sqlite_client <- create_mcp_client(
  command = "npx",
  args = c("-y", "@modelcontextprotocol/server-sqlite", "--db", "my_data.db")
)

# Handshake is performed automatically. Check if alive:
sqlite_client$is_alive()
```

### 2. Discovering Tools

Once connected, you can list the tools provided by the server.

```{r}
tools <- sqlite_client$list_tools()
# Each tool has a name, description, and JSON Schema input parameters.
```

### 3. Integrating with aisdk Tools

To use these tools with `generate_text` or an `Agent`, you must convert them into SDK-compatible `Tool` objects using `$as_sdk_tools()`.

```{r}
sdk_tools <- sqlite_client$as_sdk_tools()

model <- create_openai()$language_model("gpt-4o")

# Run a query using the MCP tools
response <- generate_text(
  model = model,
  prompt = "Query the 'iris' table and tell me the average Sepal.Length.",
  tools = sdk_tools
)

cat(response$text)
```

## Advanced Usage

### Resources and Notifications
The `McpClient` also supports reading **Resources** (static data/docs) and sending **Notifications** to the server.

```{r}
# List resources
resources <- sqlite_client$list_resources()

# Read a resource URI
# data <- sqlite_client$read_resource("sqlite://main/table/iris/schema")
```

### Cleanup
Always close the client connection to terminate the server process.

```{r}
sqlite_client$close()
```

## Community MCP Servers
You can find hundreds of ready-to-use MCP servers in the [MCP Hub](https://mcphub.io) or the official [MCP GitHub repository](https://github.com/modelcontextprotocol/servers).
