---
title: "Chat Sessions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chat Sessions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

A `ChatSession` is the central object for managing stateful conversations. It handles message history, tool execution state, model switching, and serialization.

## Creating a Session

You can create a session directly or through an agent.

```{r}
library(aisdk)

model <- create_openai()$language_model("gpt-4o")

# Option 1: Direct creation
session <- ChatSession$new(
  model = model,
  system_prompt = "You are a helpful R expert."
)

# Option 2: From an agent (inherits agent's tools and persona)
# coder <- create_coder_agent()
# session <- coder$create_session(model = model)
```

## Basic Interaction

Use `$send()` for standard requests and `$send_stream()` for streaming responses.

```{r}
# Send a message
response <- session$send("How do I use R6 classes?")
cat(response$text)

# The session automatically remembers the history
response2 <- session$send("Can you give me a code example?")
cat(response2$text)
```

## Model Switching

You can switch the underlying model at any point. The conversation history remains intact.

```{r}
# Switch to a different model mid-session
session$switch_model("anthropic:claude-3-5-sonnet-20241022")

# Continue the conversation with the new model
session$send("Now rewrite that example using S4.")
```

## State Management

Sessions provide a shared environment and a "memory" list, which are essential for multi-agent systems where agents need to share data.

### Shared Environment (`.envir`)
This is a real R environment where agents can store and retrieve data frames, models, and other objects.
```{r}
env <- session$get_envir()
env$data <- iris

# Agents with tool access can now see 'data'
```

### Shared Memory
A simple key-value store for metadata or state flags.
```{r}
session$set_memory("current_status", "processing_data")
status <- session$get_memory("current_status")
```

## Persistence

Sessions can be saved to disk as RDS or JSON files and restored later.

```{r}
# Save session state
session$save("my_analysis.json")

# Restore session in a new process
new_session <- ChatSession$new(model = model)
new_session$restore("my_analysis.json")
```

## Console Chat

For a quick interactive REPL experience in the R console, use `console_chat()`.
It is built on top of the same session model, so you can chat interactively,
switch models, inspect history, and persist state without leaving the terminal.

```{r}
# Starts an interactive chat loop in the console
# console_chat(model = "openai:gpt-4o")
```

Key capabilities:

- `clean`, `inspect`, and `debug` view modes for different levels of output detail
- inline tool execution summaries for agent-driven turns
- an inspector overlay for the latest turn or a specific tool
- session persistence with `/save` and `/load`
- model switching with `/model <id>`
- optional local execution mode with `/local [on|off]`

Useful commands during a session:

- `/inspect on`: switch to inspect mode
- `/inspect turn`: open the latest turn in the inspector overlay
- `/inspect tool <index>`: inspect a specific tool
- `/inspect next` / `/inspect prev`: move between tools in the inspector
- `/debug on`: enable detailed troubleshooting output
- `/stats`: review session usage statistics
- `/history`: show the conversation history

For a deeper terminal walkthrough, including inspect/debug modes and overlay
navigation, see the [Console Chat](console_chat.html) vignette. For more
details on how agents interact with sessions, see the [Agents](agents.html)
vignette.
