---
title: "Skill System"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Skill System}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The **Skill System** in `aisdk` implements a **Progressive Disclosure** pattern. This allows agents to access vast amounts of specialized knowledge and code without overwhelming the LLM's context window.

## The 3-Level Disclosure Pattern

1.  **Level 1 (Metadata)**: Only the skill name and a brief description are loaded initially. This is injected into the character's system prompt.
2.  **Level 2 (Instructions)**: If the agent decides a skill is relevant, it calls `load_skill()`. This returns the full `SKILL.md` content (detailed guidelines, examples, etc.).
3.  **Level 3 (Implementation)**: The agent can then browse and execute specific R scripts within the skill's `scripts/` directory or read reference files in `assets/`.

## Skill Structure

A skill is a directory (often stored in a global `skills/` folder) with the following structure:

```text
my_skill/
├── SKILL.md       # Required: Metadata (YAML) + Instructions (Markdown)
├── scripts/       # Optional: R scripts for the agent to run
│   └── analyze.R
└── assets/        # Optional: Reference data, schemas, or templates
    └── template.json
```

### Example: `data_cleaner/SKILL.md`

```markdown
---
name: data_cleaner
description: Specialized tools for cleaning and prehistoric R data frames.
---
# Data Cleaning Guidelines
- Always check for NAs before summarizing.
- Use `clean_names.R` to normalize column headers.
```

## Using Skills with Agents

Agents can automatically load skills from a directory.

```{r}
library(aisdk)

# Create an agent that 'auto-discovers' skills in a specific path
agent <- create_agent(
  name = "Analyst",
  description = "Data analyst with specialized skills",
  skills = "./my_project/skills" # Path to your skills directory
)

# When running, the agent will see 'data_cleaner' in its available tools list.
model <- create_openai()$language_model("gpt-4o")
result <- agent$run("Clean the 'dirty_data' frame in the session.", model = model)
```

## Manual Skill Management

You can also use a `SkillRegistry` for more fine-grained control.

```{r}
# 1. Create a registry
registry <- create_skill_registry("./skills")

# 2. Add skills to a model interaction manually
response <- generate_text(
  model = model,
  prompt = "Use the 'genomics' skill to identify markers.",
  skills = registry
)
```

## Skill Tools

When an agent or `generate_text` is given a skill registry, the following tools are automatically injected:

- `load_skill(skill_name)`: Loads the Level 2 instructions.
- `list_skill_scripts(skill_name)`: Lists Level 3 scripts.
- `execute_skill_script(skill_name, script_name, ...)`: Runs an R script.
- `read_skill_resource(skill_name, resource_name)`: Reads an asset file.

This layered approach ensures that the LLM only "pays" for the information it actually needs to solve the current task.
