---
title: "Structured Outputs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structured Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

In many AI applications, you need the model to return data in a specific, machine-readable format (like JSON) rather than free-form text. **aisdk** provides robust support for structured outputs using JSON Schema.

## Using `generate_object` with `schema`

The primary way to get structured data is to call `generate_object()` and pass a `z_schema` to the `schema` argument.

```{r}
library(aisdk)

# 1. Define your data structure
recipe_schema <- z_object(
  name = z_string("Name of the dish"),
  ingredients = z_array(z_string(), "List of ingredients"),
  prep_time_mins = z_integer("Preparation time in minutes")
)

# 2. Request structured output
model <- create_openai()$language_model("gpt-4o")

result <- generate_object(
  model = model,
  prompt = "Give me a simple recipe for a classic Omelette.",
  schema = recipe_schema
)

# The result$object contains the parsed R list
print(result$object$name)
print(result$object$ingredients)
```

## Extracting Tabular Data

The `z_dataframe()` helper is specifically designed for extracting lists of objects that can be easily converted into an R `data.frame`.

```{r}
# Define a schema for a list of genes
gene_schema <- z_dataframe(
  gene_symbol = z_string("Standard gene symbol"),
  fold_change = z_number("Log2 fold change value"),
  p_value = z_number("Adjusted p-value")
)

result <- generate_object(
  model = model,
  prompt = "Extract gene expression data: BRCA1 (FC=2.5, p=0.01), TP53 (FC=-1.2, p=0.04).",
  schema = gene_schema
)

# Convert to a tidy data frame
library(dplyr)
df <- bind_rows(result$object)
print(df)
```

## How it Works

When you provide a `schema` to `generate_object()`, **aisdk** intelligently determines the best way to interact with the LLM API:
1.  **Native Format Constraints**: For providers that support it (like OpenAI and Anthropic), it utilizes their native "JSON Mode" or strict "Structured Outputs" feature to guarantee the output matches the schema.
2.  **Prompt Injection & Repair**: For providers that lack native JSON schema support (like Stepfun or some Volcengine models), `aisdk` gracefully falls back. It automatically appends the formalized JSON Schema into the system prompt, requests JSON output, and performs post-generation validation and automatic structural repair.

## Nested Structures

The Schema DSL supports arbitrary nesting, allowing you to model complex hierarchical data.

```{r}
report_schema <- z_object(
  metadata = z_object(
    author = z_string(),
    timestamp = z_string()
  ),
  sections = z_array(
    z_object(
      title = z_string(),
      content = z_string(),
      word_count = z_integer()
    )
  )
)
```

For a detailed list of all available schema helpers, see the [Tools](tools.html) vignette.
