---
title: "Using Tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Tools allow AI models to perform actions, like calculating numbers, fetching data, or interacting with the filesystem. **aisdk** makes it easy to turn R functions into tools.

## Defining Tools

You define a tool using the `tool()` function. You also need to define a schema for the tool's expected arguments using helper functions (`z_object`, `z_string`, `z_number`, etc.).

### Example: Weather Tool

```{r}
library(aisdk)

# 1. Define the input schema
weather_schema <- z_object(
  location = z_string(description = "The city name, e.g., 'Tokyo'"),
  unit = z_enum(c("celsius", "fahrenheit"), description = "Temperature unit")
)

# 2. Create the tool
get_weather <- tool(
  name = "get_weather",
  description = "Get the current weather for a city",
  parameters = weather_schema,
  execute = function(args) {
    # In a real app, call an API here
    # For demo, returning mock string
    location <- args$location
    unit <- args$unit %||% "celsius"
    
    paste0("Weather in ", location, ": 22°C (", unit, ")")
  }
)
```

## Using Tools with Models

To use tools, pass them as a list to `generate_text()`.

```{r}
model <- create_openai()$language_model("gpt-4o")

result <- generate_text(
  model = model,
  prompt = "What's the weather like in Paris?",
  tools = list(get_weather)
)

# If the model decided to call the tool, the result text will contain the final answer
# after the tool was executed and the result fed back to the model.
cat(result$text)
```

## Manual Execution

You can also run tools manually to test them:

```{r}
result <- get_weather$run(list(location = "London", unit = "celsius"))
print(result)
```

## Supported Schema Types

| Helper | Type | Description |
|--------|------|-------------|
| `z_string()` | String | Text inputs |
| `z_number()` | Number | Integers or floats |
| `z_boolean()` | Boolean | TRUE/FALSE flags |
| `z_enum()` | String | Restricted set of choices |
| `z_array()` | Array | List of items |
| `z_object()` | Object | Nested structure |
| `z_dataframe()` | Array | List of objects (R data.frame) |

