---
title: "Using Tools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Tools allow AI models to perform actions like fetching real-time data, interacting with the filesystem, or executing complex calculations in R. In **aisdk**, tools are first-class objects that bridge the gap between LLM function calling and R functions.

## Defining Tools

A tool consists of a `name`, a `description` (for the LLM), a `parameters` schema, and an `execute` function.

### The Schema DSL

**aisdk** provides a Zod-like DSL for defining JSON Schema structure in R. This ensures the LLM provides valid arguments and enables automatic documentation generation.

| Helper | JSON Type | R Type |
|--------|-----------|--------|
| `z_string()` | `string` | `character` |
| `z_number()` | `number` | `numeric` |
| `z_integer()` | `integer` | `integer` |
| `z_boolean()` | `boolean` | `logical` |
| `z_enum()` | `string` | `character` (subset) |
| `z_object()` | `object` | `list` |
| `z_array()` | `array` | `list` |
| `z_dataframe()`| `array` | `list` of `objects` |

### Example: A Scientific Calculator Tool

```{r}
library(aisdk)

calc_tool <- tool(
  name = "calculate",
  description = "Perform a mathematical calculation on two numbers",
  parameters = z_object(
    a = z_number("The first number"),
    b = z_number("The second number"),
    operation = z_enum(c("add", "subtract", "multiply", "divide"), "The operator")
  ),
  execute = function(args) {
    a <- args$a
    b <- args$b
    op <- args$operation

    switch(op,
      add = a + b,
      subtract = a - b,
      multiply = a * b,
      divide = if (b != 0) a / b else "Error: Division by zero"
    )
  }
)
```

## Using Tools

You can use tools with `generate_text()` or with an `Agent`.

### With `generate_text()`

```{r}
model <- create_openai()$language_model("gpt-4o")

result <- generate_text(
  model = model,
  prompt = "What is 123 multiplied by 456?",
  tools = list(calc_tool)
)

cat(result$text)
# > "The result of 123 multiplied by 456 is 56088."
```

## The Execution Environment

Inside the `execute` function, you can access the current session environment via `args$.envir`. This allows tools to share state (like data frames or models) across multiple steps.

```{r}
state_tool <- tool(
  name = "save_state",
  description = "Save a value to the session environment",
  parameters = z_object(
    key = z_string("Variable name"),
    value = z_any("Value to save")
  ),
  execute = function(args) {
    # .envir is automatically injected by the SDK
    env <- args$.envir
    if (!is.null(env)) {
      env[[args$key]] <- args$value
      return(paste("Saved", args$key, "to session."))
    }
    "Error: No session environment found."
  }
)
```

## Advanced Patterns

### Tool Layers
You can categorize tools into "layers" for better organization.
```{r}
my_tool <- tool(..., layer = "data_analysis")
```

### Read-Only vs. State-Mutating

When building autonomous agents, it is a best practice to strictly separate tools by their side effects:

- **Read-Only**: Tools that fetch data (e.g., `read_file`, `query_database`). These can typically be executed automatically.
- **State-Mutating**: Tools that change the environment or external systems (e.g., `write_file`, `execute_query`). 

You can use the `aisdk` **Permission Hooks** and **Human-in-the-Loop (HITL)** system to automatically pause execution and require explicit user approval whenever an agent attempts to call a state-mutating tool.

### Automatic Repair
**aisdk** includes built-in mechanisms to repair small JSON errors in tool arguments and fuzzy-match tool names if the LLM makes a minor typo.

### Computer Abstraction
The `create_computer_tools()` function provides a standardized set of tools for filesystem and bash operations, complete with safety sandboxing.

```{r}
computer_tools <- create_computer_tools(working_dir = getwd())
# Includes: bash, read_file, write_file, execute_r_code
```

For more details on computer tools, see the [Computer Abstraction](computer_abstraction.html) vignette.

